Model based reinforcement learning at FERMI FEL

### 2020-11-17 main.tex
txs auto checkin

\begin{abstract}
In this paper we discuss a model based reinforcement learning approach in comparison to a model free reinforcement learning approach applied at the FERMI FEL system. Both algorithms and approaches are new in this context and the main purpose of this paper is to show that reinforcement learning can be applied on an operational level in a feasible training time on real accelerator physics problems. In terms of sample-complexity the model-based approach is faster, while the finale performance of the model free method is superior - the so called asymptotic performance. The model-based algorithm is done in a Dyna-style using an uncertainty aware model and the model-free algorithm is based on tailored deep Q-learning using some tricks to increase the sample efficiency
\end{abstract}
%\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle
\begin{itemize}
	\item Noise benchmarks with pendulum
	\item Benchmark NAF2 with pendulum
	\item What is our improvement concerning the algorithm
	\item The first time model based on accelerator problem, 
\end{itemize}
%\tableofcontents
\section{Introduction and Motivation}
In particle accelerators one main goal is to provide stable and reproducible performance. In order to achieve this, a number of control problems have to be considered simultaneously. Especially if there is no way to model the physics, one might use optimisation techniques as e.g. derivative free optimisers (DFOs) or model-based optimisations as Gaussian processes. Another promising way is to apply reinforcement learning which has several advantages over optimisation methods:
\begin{itemize}
    \item Covers a larger class of problems.
    \item Solves the problem not always completely from the beginning as an DFO.
    \item Ability to use existing data.
    \item The underlying structure of the problem might be deduced.
\end{itemize}
RL optimises a sequence of decisions. On critical aspect using RL is the number of iteration to train a controller.
In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at Elettra Sincrotrone Trieste. Three algorithms were applied successfully showing different advantages, which will e discussed in the following.

@@

% \bibliography{bibliography}
\addcontentsline{toc}{chapter}{Bibliography}

\begin{thebibliography}{9}

\bibitem{lsa} D. Jacquet, R. Gorbonosov, G. Kruk, "LSA - The High Level Application Software of the LHC and its Performance during the first 3 years of Operation", ICALEPS, San Francisco, CA, USA, 6 - 11 Oct 2013, pp.thppc058.

\bibitem{multiobjective-optimization} A. Edelen, N. Neveu, M. Frey, Y. Huber, C. Mayes, and A. Adelmann, "Machine learning for orders of magnitude speedup in multiobjective optimization of particle accelerator systems", Phys. Rev. Accel. Beams 23, 044601, 2020.

\bibitem{collimator-alignment} G. Azzopardi, A. Muscat, G. Valentino, S. Redaelli, B. Salvachua, "Operational results of LHC collimator alignment using machine learning", Proc. IPAC'19, Melbourne, Australia, pp. 1208-1211, 2019.

\bibitem{SPSAlignment} %InProceedings (Hirlaender2019)
 S. Hirlaender, M. Fraser, B. Goddard , V. Kain, J. Prieto, L. Stoel, M. Szakaly, F. Velotti, "Automatisation of the SPS ElectroStatic Septa Alignment",  
in 10th Int. Partile Accelerator Conf.(IPAC'19), 2019, p. 4001-4004.

\bibitem{liu} E. Shaposhnikova et al., "LHC Injectors Upgrade (LIU) Project at CERN", 7$^{th}$ International Particle Accelerator Conference, Busan, Korea, 8 - 13 May 2016, pp.MOPOY059, \url{https://doi.org/10.18429/JACoW-IPAC2016-MOPOY059}.

\bibitem{awake} E. Adli, A. Ahuja, O. Apsimon, R. Apsimon, A.-M. Bachmann, D. Barrientos, F. Batsch, J. Bauche, V. B. Olsen, M. Bernardini \textit{et al.}, “Acceleration of electrons in the plasma wakefield of a proton bunch,” Nature 561, 363–367 (2018), \url{https://doi.org/10.1038/s41586-018-0485-4)}.

\bibitem{scheinker} A. Scheinker \textit{et al.}, "Online multi-objective particle accelerator optimization of the AWAKE electron beam line for simultaneous emittance and orbit control", AIP Advances 10, 055320 (2020), \url{https://doi.org/10.1063/5.0003423}.

\bibitem{barto-sutton} R. Sutton, A. Barto, "Introduction to Reinforcement Learning",  MIT Press, Cambridge, MA, USA, 2018.

\bibitem{LQR} H. Kwakernaak, R. Sivan, "Linear Optimal Control Systems", Wiley-Interscience, 1972.

\bibitem{SBP} A. Nagabandi, G. Kahn, R. S. Fearing, S. Levine, "Neural Network Dynamics for Model-Based Deep Reinforcement Learning with Model-Free Fine-Tuning", arXiv:1708.02596, 2017.

\bibitem{DYNA} R. Sutton, "Dyna, an integrated architecture for learning, planning, and reacting", AAAI Spring Symposium, pp. 151-155, 1991, \url{https://doi.org/10.1145/122344.122377}.

\bibitem{benchmarkmbrl} T. Wang et al., “Benchmarking Model-Based Reinforcement Learning”,  \url{https://arxiv.org/abs/1907.02057v1}, 2019-07-03.

\bibitem{dqn} V. Mnih \textit{et al.}, "Human-level control through deep reinforcement learning", Nature 518, 529–533 (2015).

\bibitem{ddpg} T. Lillicrap~\textit{et al.}, "Continuous control with deep reinforcement learning", Proc. ICLR 2016 \url{https://arxiv.org/abs/1509.02971}.

\bibitem{naf} S. Gu, T. Lillicrap, I. Sutskever, S. Levine, "Continuous Deep Q-Learning with Model-based Acceleration", in Proc. 33rd International Conference on Machine Learning, New York, NY, USA, 2016.

\bibitem{td3} S. Fujimoto, H. van Hoof, D. Meger, "Addressing Function Approximation Error in Actor-Critic Methods", arXiv:1802.09477, 2018.

\bibitem{deeppilco} Y. Gal, R. McAllister, and C. E. Rasmussen, “Improving PILCO with Bayesian neural network dynamics models”, in Data-Efficient Machine Learning workshop, ICML, 2016, vol. 4, p. 34.

\bibitem{per} T. Schaul, J. Quan, I. Antonoglou, D. Silver, "Prioritized Experience Replay", arXiv:1511.05952, 2015-11-18.

\bibitem{pernafgit} S. Hirlaender, \textit{PER-NAF} available at \url{https://github.com/MathPhysSim/PER-NAF/} and
\url{https://pypi.org/project/pernaf/}.
\bibitem{pyjapc} "pyjapc" available at \url{https://pypi.org/project/pyjapc/}.

\bibitem{openai_gym} \url{http://gym.openai.com}

\bibitem{madx} MAD-X documentation and source code available at \url{https://mad.web.cern.ch/mad/}.

\bibitem{linac4} G. Bellodi, "Linac4 Commissioning Status and Challenges to Nominal Operation", 61$^st$ ICFA Advanced Beam Dynamics Workshop on High-Intensity and High-Brightness Hadron Beams, Daejeon, Korea, 17 - 22 Jun 2018, pp.MOA1PL03, \url{https://doi.org/10.18429/JACoW-HB2018-MOA1PL03}.

\bibitem{auto_matching} F. Velotti, B. Goddard et al., "Automatic AWAKE electron beamline setup using unsupervised machine learning", to be submitted to Phys. Rev. Accel. Beams.

\bibitem{Kumar_1} A. Kumar, A. Gupta, S. Levine, “DisCor: Corrective Feedback in Reinforcement Learning via Distribution Correction,” arXiv:2003.07305, 2020-03-16.

\bibitem{Kumar_2} A. Kumar, J. Fu, G. Tucker, S. Levine, “Stabilizing Off-Policy Q-Learning via Bootstrapping Error Reduction,” arXiv:1906.00949, 2019-06-03.

% NEW REFERENCES
\bibitem[bruchon2020basic]{Bruchon, Niky, et al. "Basic Reinforcement Learning Techniques to Control the Intensity of a Seeded Free-Electron Laser." Electronics 9.5 (2020): 781.}
% https://www.mdpi.com/2079-9292/9/5/781/htm

\end{thebibliography}

>>>>

\begin{abstract}
In this paper we discuss a model based reinforcement learning approach in comparison to a model free reinforcement learning approach applied at the FERMI FEL system. Both algorithms and approaches are new in this context and the main purpose of this paper is to show that reinforcement learning can be applied on an operational level in a feasible training time on real accelerator physics problems. In terms of sample-complexity the model-based approach is faster, while the finale performance of the model free method is superior - the so called asymptotic performance. The model-based algorithm is done in a Dyna-style using an uncertainty aware model and the model-free algorithm is based on tailored deep Q-learning using some tricks to increase the sample efficiency
\end{abstract}
%\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle
\begin{itemize}
	\item Noise benchmarks with pendulum
	\item Benchmark NAF2 with pendulum
	\item What is our improvement concerning the algorithm
	\item The first time model based on accelerator problem compared to a novel state of the art MFRL: NAF2
	\item Tailored to accelerators - short horizons are standard?
	\item Improvements of the algorithms:
	\begin{itemize}
		\item Adding the noise feature
		\item Jumping feature using SAC
		\item Noise in the policy
		\item discuss different approaches to e.g. dynamic waiting time
	\end{itemize}
\end{itemize}
%\tableofcontents
\section{Introduction and Motivation}
In particle accelerators one main goal is to provide stable and reproducible performance. In order to achieve this, a number of control problems have to be considered simultaneously. Especially if there is no way to model the physics, one might use optimisation techniques as e.g. derivative free optimisers (DFOs) or model-based optimisations as Gaussian processes. Another promising way is to apply reinforcement learning which has several advantages over optimisation methods:
\begin{itemize}
    \item Covers a larger class of problems.
    \item Solves the problem not always completely from the beginning as an DFO.
    \item Ability to use existing data.
    \item The underlying structure of the problem might be deduced.
\end{itemize}
RL optimises a sequence of decisions. On critical aspect using RL is the number of iteration to train a controller.
In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at Elettra Sincrotrone Trieste. Three algorithms were applied successfully showing different advantages, which will e discussed in the following.

@@

% \bibliography{bibliography}
%\addcontentsline{toc}{chapter}{Bibliography}
%
%\begin{thebibliography}{9}
%
%\bibitem{lsa} D. Jacquet, R. Gorbonosov, G. Kruk, "LSA - The High Level Application Software of the LHC and its Performance during the first 3 years of Operation", ICALEPS, San Francisco, CA, USA, 6 - 11 Oct 2013, pp.thppc058.
%
%\bibitem{multiobjective-optimization} A. Edelen, N. Neveu, M. Frey, Y. Huber, C. Mayes, and A. Adelmann, "Machine learning for orders of magnitude speedup in multiobjective optimization of particle accelerator systems", Phys. Rev. Accel. Beams 23, 044601, 2020.
%
%\bibitem{collimator-alignment} G. Azzopardi, A. Muscat, G. Valentino, S. Redaelli, B. Salvachua, "Operational results of LHC collimator alignment using machine learning", Proc. IPAC'19, Melbourne, Australia, pp. 1208-1211, 2019.
%
%\bibitem{SPSAlignment} %InProceedings (Hirlaender2019)
% S. Hirlaender, M. Fraser, B. Goddard , V. Kain, J. Prieto, L. Stoel, M. Szakaly, F. Velotti, "Automatisation of the SPS ElectroStatic Septa Alignment",  
%in 10th Int. Partile Accelerator Conf.(IPAC'19), 2019, p. 4001-4004.
%
%\bibitem{liu} E. Shaposhnikova et al., "LHC Injectors Upgrade (LIU) Project at CERN", 7$^{th}$ International Particle Accelerator Conference, Busan, Korea, 8 - 13 May 2016, pp.MOPOY059, \url{https://doi.org/10.18429/JACoW-IPAC2016-MOPOY059}.
%
%\bibitem{awake} E. Adli, A. Ahuja, O. Apsimon, R. Apsimon, A.-M. Bachmann, D. Barrientos, F. Batsch, J. Bauche, V. B. Olsen, M. Bernardini \textit{et al.}, “Acceleration of electrons in the plasma wakefield of a proton bunch,” Nature 561, 363–367 (2018), \url{https://doi.org/10.1038/s41586-018-0485-4)}.
%
%\bibitem{scheinker} A. Scheinker \textit{et al.}, "Online multi-objective particle accelerator optimization of the AWAKE electron beam line for simultaneous emittance and orbit control", AIP Advances 10, 055320 (2020), \url{https://doi.org/10.1063/5.0003423}.
%
%\bibitem{barto-sutton} R. Sutton, A. Barto, "Introduction to Reinforcement Learning",  MIT Press, Cambridge, MA, USA, 2018.
%
%\bibitem{LQR} H. Kwakernaak, R. Sivan, "Linear Optimal Control Systems", Wiley-Interscience, 1972.
%
%\bibitem{SBP} A. Nagabandi, G. Kahn, R. S. Fearing, S. Levine, "Neural Network Dynamics for Model-Based Deep Reinforcement Learning with Model-Free Fine-Tuning", arXiv:1708.02596, 2017. 
%
%\bibitem{DYNA} R. Sutton, "Dyna, an integrated architecture for learning, planning, and reacting", AAAI Spring Symposium, pp. 151-155, 1991, \url{https://doi.org/10.1145/122344.122377}.
%
%\bibitem{benchmarkmbrl} T. Wang et al., “Benchmarking Model-Based Reinforcement Learning”,  \url{https://arxiv.org/abs/1907.02057v1}, 2019-07-03.
%
%\bibitem{dqn} V. Mnih \textit{et al.}, "Human-level control through deep reinforcement learning", Nature 518, 529–533 (2015).
%
%\bibitem{ddpg} T. Lillicrap~\textit{et al.}, "Continuous control with deep reinforcement learning", Proc. ICLR 2016 \url{https://arxiv.org/abs/1509.02971}.
%
%\bibitem{naf} S. Gu, T. Lillicrap, I. Sutskever, S. Levine, "Continuous Deep Q-Learning with Model-based Acceleration", in Proc. 33rd International Conference on Machine Learning, New York, NY, USA, 2016.
%
%\bibitem{td3} S. Fujimoto, H. van Hoof, D. Meger, "Addressing Function Approximation Error in Actor-Critic Methods", arXiv:1802.09477, 2018.
%
%\bibitem{deeppilco} Y. Gal, R. McAllister, and C. E. Rasmussen, “Improving PILCO with Bayesian neural network dynamics models”, in Data-Efficient Machine Learning workshop, ICML, 2016, vol. 4, p. 34.
%
%\bibitem{per} T. Schaul, J. Quan, I. Antonoglou, D. Silver, "Prioritized Experience Replay", arXiv:1511.05952, 2015-11-18.
%
%\bibitem{pernafgit} S. Hirlaender, \textit{PER-NAF} available at \url{https://github.com/MathPhysSim/PER-NAF/} and
%\url{https://pypi.org/project/pernaf/}.
%\bibitem{pyjapc} "pyjapc" available at \url{https://pypi.org/project/pyjapc/}.
%
%\bibitem{openai_gym} \url{http://gym.openai.com}
%
%\bibitem{madx} MAD-X documentation and source code available at \url{https://mad.web.cern.ch/mad/}.
%
%\bibitem{linac4} G. Bellodi, "Linac4 Commissioning Status and Challenges to Nominal Operation", 61$^st$ ICFA Advanced Beam Dynamics Workshop on High-Intensity and High-Brightness Hadron Beams, Daejeon, Korea, 17 - 22 Jun 2018, pp.MOA1PL03, \url{https://doi.org/10.18429/JACoW-HB2018-MOA1PL03}.
%
%\bibitem{auto_matching} F. Velotti, B. Goddard et al., "Automatic AWAKE electron beamline setup using unsupervised machine learning", to be submitted to Phys. Rev. Accel. Beams.
%
%\bibitem{Kumar_1} A. Kumar, A. Gupta, S. Levine, “DisCor: Corrective Feedback in Reinforcement Learning via Distribution Correction,” arXiv:2003.07305, 2020-03-16.
%
%\bibitem{Kumar_2} A. Kumar, J. Fu, G. Tucker, S. Levine, “Stabilizing Off-Policy Q-Learning via Bootstrapping Error Reduction,” arXiv:1906.00949, 2019-06-03.
%
%% NEW REFERENCES
%\bibitem[bruchon2020basic]{Bruchon, Niky, et al. "Basic Reinforcement Learning Techniques to Control the Intensity of a Seeded Free-Electron Laser." Electronics 9.5 (2020): 781.}
%% https://www.mdpi.com/2079-9292/9/5/781/htm
%
%
%
%\end{thebibliography}
